Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model

### 2025-10-30 analysis/paper/access-ice.tex
Renders

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-mathphys-num,
  referee]{sn-jnl}

@@

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Model},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew Wheeler; Phillip Reid},
  pdfkeywords={sea ice, predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Evaluating the Importance of Initial Conditions for Antarctic Sea
Ice Seasonal Predictability with a Fully Coupled Model]{Evaluating the
Importance of Initial Conditions for Antarctic Sea Ice Seasonal
Predictability with a Fully Coupled Model}

% author setup
\author*[1,2]{\fnm{Elio} \sur{Campitelli}}\email{elio.campitelli@monash.edu}\author[1,2]{\fnm{Ariaan} \sur{Purich}}\author[1,2]{\fnm{Julie} \sur{Arblaster}}\author[3]{\fnm{Eun-Pa} \sur{Lim}}\author[3]{\fnm{Matthew} \sur{Wheeler}}\author[3]{\fnm{Phillip} \sur{Reid}}
% affil setup
\affil[1]{\orgdiv{School of Earth, Atmosphere and
Environment}, \orgname{Monash University, Australia,}}
\affil[2]{\orgdiv{Securing Antarctica's Environmental
Future}, \orgname{Monash University, Australia}}
\affil[3]{\orgdiv{Research, Bureau of Meteorology}}

@@

\abstract{Accurate Antarctic sea ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. The two
versions of the Australian Bureau of Meteorology's ACCESS seasonal
forecast system, ACCESS-S1 and ACCESS-S2, use identical model
configuration and differ only in their initial conditions; primarily in
that ACCESS-S2 does not assimilate sea-ice observations. This makes for
a good ``natural experiment'' to analyse the role of initial conditions
on Antarctic sea ice forecasts using a long run and a moderately large
ensemble with a fully coupled model. Our analysis reveals that both
systems experience an extended melt season and delayed growth phase.
This leads to a significant negative sea-ice extent bias which is
corrected only in ACCESS-S1 by the data assimilation system. The impact
of the initial conditions on forecast errors varies dramatically by
season: summer and autumn initial conditions (January-April) provide
predictive skill for up to three months, with February initial
conditions being particularly crucial. In contrast, winter forecasts of
the two systems are statistically indistinguishable after only two
weeks. Regional analysis of forecast skill suggests that this winter
predictability barrier is most dramatic over Eastern Antarctica, where
even ACCESS-S1 shows negative skill. These findings highlight the
critical importance of comprehensive year-round sampling in
predictability studies and suggest that operational sea-ice data
assimilation efforts should prioritize the summer-autumn period when
initial conditions have maximum impact on forecast skill.}

% keywords
\keywords{sea ice,  predictability,  initial conditions,  forecasting}

@@

However, Antarctic sea ice forecasts have lagged behind Arctic sea ice
forecasts due to model biases, inherent large variability and complexity
\citep{zampieri2019, gao2024a}. Dynamical seasonal forecasts of summer
Antarctic sea ice have been shown to perform worse than relatively
simpler statistical methods \citep{massonnet2023} and machine learning
approaches (e.g. \citet{dong2024}, \citet{lin2025}), which also
underscores the need for better understanding of sea-ice dynamics, and
drivers of its variability.

Good initial conditions are generally required for a good forecast,
however, it is not entirely known to what extent accurate sea ice
initial conditions affect the quality of the forecast and at what
timescales. Exploring seasonal predictions of Arctic sea ice,
\citet{guemas2016} found that sea ice initial conditions are important
in autumn to predict summer sea ice, but the impact wasn't as dramatic
when predicting winter sea ice. \citet{day2014} also found
seasonally-varying differences in the effect of initialisation, noting
that accurate Arctic sea-ice thickness lead to improved sea ice
forecasts initialised in July but not when initialised in January.

For the Antarctic, \citet{holland2013} studied the initial-value
predictability of Antarctic sea ice in a perfect model study using the
CCSM3 model. They found that sea-ice and ocean initial conditions
provide predictive information to forecast sea-ice edge location several
months in advance and that some predictability is retained for up to two
years thanks to ocean heat content anomalies that are advected eastward.
This is in contrast with \citet{marchi2020}, who ran perfect model
experiments to argue that uncertainty in the predicted atmospheric state
and evolution is the main driver of uncertainty in Antarctic sea-ice
extent prediction on seasonal timescales, with sea-ice and ocean initial
conditions having lesser importance. More recently, \citet{morioka2022}
studied decadal forecasts of Antarctic sea ice and found that
initialising ocean and sea ice improved the correlation between
simulated and observed sea-ice concentration evolution in the
Amundsen--Bellingshausen Sea. It is hard to compare these studies since
they are based on forecasts initialised at different times of the year
and different frameworks: \citet{holland2013} ran 20 ensemble members
initialised on the 1st of January of a particular year,
\citet{marchi2020} ran forecasts from the 1st of March and 1st of
September and \citet{morioka2022} run forecasts only from the 1st of
March. \citet{marchi2020} also used a coupled ocean--sea-ice model
instead of a fully coupled model like \citet{holland2013} did.
\citet{morioka2022} used observed sea ice initial conditions and
compared with observations, while \citet{marchi2020} and
\citet{holland2013} are perfect model studies.

@@

Since model configuration is identical between ACCESS-S1 and ACCESS-S2,
they form a sort of ``natural experiment'' where the same forecasting
model was run over a long period of time with multiple ensemble
forecasts initialised throughout the year, with the only difference
being the initial conditions. This provides an opportunity to test the
effect of sea ice initial conditions on the forecast of sea-ice
concentrations and the climate.

In this study we compare sea ice hindcasts produced by ACCESS-S1 and
ACCESS-S2. We focus on seasonality of errors and biases and the effect
of the data assimilation system. This comparison will also inform future
work with the prediction system as a research tool to better understand
the dynamics and variability of the Antarctic sea ice and its impacts on
the climate system as well as to explore the potential of using its sea
ice forecasts for decision-making. The work will also serve as a
benchmark for future prediction systems to attempt to improve upon.

@@

Both systems take atmospheric initial conditions derived from
ERA-interim \citep{dee2011} for their hindcasts. The main difference
between the hindcasts of the two systems are the ocean and sea ice
initial conditions. ACCESS-S1's ocean and sea ice initial conditions
come from the Met Office FOAM system, which uses a multivariate,
incremental three-dimensional variational (3D-Var),
first-guess-at-appropriate-time (FGAT) data assimilation scheme
\citep{waters2015} and assimilates sea surface temperature (SST), sea
surface height (SSH), in situ temperature and salinity profiles, and
satellite observations of sea-ice concentration using the EUMETSAT
OSISAF product described in the next section. ACCESS-S2, on the other
hand, is initialised from ocean conditions generated by the BoM weakly
coupled ensemble data assimilation scheme described in \citet{wedd2022}.
This scheme uses an optimal interpolation method and assimilates
temperature and salinity profiles from EN4 \citep{good2013}. SSTs are
nudged to Reynolds OISSTv2.1 \citep{reynolds2007} in areas where SSTs
are over 0ºC and Sea Surface Salinity is weakly nudged to the World
Ocean Atlas 2013 climatology \citep{zweng2013}.

Of most relevance for this work, sea-ice concentrations are not
assimilated in ACCESS-S2. Assimilation cycles are performed daily. The
coupled model runs for 24 hours initialised from the previous cycle.
Then the restart file fields of the ocean component are used as first
guess in the data assimilation cycle and the innovations are used to
build the next ocean initial conditions for the following cycle. The
atmosphere fields from that daily integration are not used and instead
the model atmosphere is initialised using ERA-Interim. The sea ice
initial conditions for the next cycle are the unaltered output of the
previous daily integration. Then the cycle starts again and the coupled
model runs for another 24 hours. During this integration the sea-ice
component is affected by the ocean innovations and the new atmosphere
initial conditions via the coupler.

The ACCESS-S1 hindcast set is made up of nine members created by
perturbing the atmospheric fields only with a random field perturbation
\citep{hudson2017} and runs for 217 days for the period 1990--2012
initialised at the first of every month. The ACCESS-S2 hindcast set used
in this study runs for the period 1981--2018. Each hindcast consists of
nine ensemble members built from three consecutive 3-member forecasts
created in the same manner as ACCESS-S1 members initialised at the first
of every month and the two previous days and run for 279 days. We
analyse the ensemble mean hindcasts unless otherwise specified.

@@

We use NOAA/NSIDC's Climate Data Record V4 {[}CDR; \citet{meier2014}{]}
as the primary sea-ice verification dataset. It takes the maximum value
of the NASA Team \citep{cavalieri1984} and NASA Bootstrap
\citep{comiso2023} sea-ice concentration products to reduce their low
concentration bias \citep{meier2014, meier2021}. Both source algorithms
use data from the Scanning Multichannel Microwave Radiometer (SMMR) on
the Nimbus-7 satellite and from the Special Sensor Microwave/Imager
(SSM/I) sensors on the Defense Meteorological Satellite Program's (DMSP)
-F8, -F11, and -F13 satellites. The data has a spatial resolution of 25
by 25 km and daily from 1978 onwards.

The European Organisation for the Exploitation of Meteorological
Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility
\citep{OSISAF} based on the SSMIS sensor is another satellite-derived
sea-ice concentration product. It is based on mostly the same sensors as
the NOAA CDR but computed independently using different algorithms.
Figures prepared with this dataset are provided in the supplementary
material and do not differ significantly from the ones prepared using
CDR.

@@

For evaluation purposes, we use a series of measures.

Sea-ice extent is defined as the area of the ocean at least 15\% covered
by sea-ice. This threshold is motivated by the limitations in satellite
retrieval, which is increasingly unreliable for lower sea-ice
concentrations \citep{cavalieri1991}.

@@

\caption{\label{fig-bias-1}Ensemble mean difference between monthly
sea-ice concentration of ACCESS-S2 ensemble mean forecast at 0-month
lead time (monthly mean values forecasted from the forecast initialised
at the first of the month) and observations.}

@@

\caption{\label{fig-bias-2}Same as Figure~\ref{fig-bias-1} but for
ACCESS-S1}

@@

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth results in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Admunsen seas.

@@

\caption{\label{fig-extent-anom}Monthly mean sea-ice extent anomalies of
the observations (black) and forecasts from ACCESS-S1 (right column;
purple) and ACCESS-S2 (left column; green) at lead times of 0, 2, 4, and
6 months. The RMSE and correlation during the overlapping period of
ACCESS-S1 and ACCESS-S2 hindcasts (1990--2013) is shown on the top left
and bottom left of each panel respectively.}

@@

Unexpectedly, for ACCESS-S2, RMSE improves with lead time, even though
the correlation degrades with lead time. This is puzzling behaviour that
goes contrary to what is usually seen in prediction models. The
explanation seems to be the mentioned increased interannual variability.
Figure~\ref{fig-extent-sd} shows the interannual standard deviation of
monthly sea-ice extent of the forecasts as a function of lead time
compared with observations. ACCESS-S1 standard deviation lies within the
observed standard deviation regardless of lead time, while ACCESS-S2
standard deviation is more than twice that of observations at zero lead
time and only approaches the observed value at 9 month lead time for
most months.

@@

As an example, Figure~\ref{fig-forecast-example} shows sea-ice
concentration anomalies (top row) and sea-ice thickness and the
difference between the two models (bottom row) for 2 May 2008
initialised one day prior; being that close to initialisation date,
these are very approximately the initial conditions. ACCESS-S1 sea-ice
concentrations anomalies are very close to observations as expected from
the system assimilating these data. ACCESS-S2 sea-ice concentration
anomalies, which are not assimilated, are not as close, but the
large-scale pattern is aligned with observations. The system simulates
large positive anomalies in the Weddell and Ross seas and slight
negative anomalies in the Amundsen and Bellingshausen seas. The fact
that ACCESS-S2 can simulate this pattern without assimilating sea-ice
data suggests that atmospheric and oceanic forcing was the dominant
factor. However, the magnitude of the sea-ice anomalies is too big. It
is plausible that this is due to the thinner ice simulated by ACCESS-S2
(bottom row).

@@

The fact that ACCESS-S1 and ACCESS-S2 share the same model configuration
and that the increased variability is more extreme at short lead times
(Fig.~\ref{fig-extent-sd}) suggests that the data assimilation procedure
is partly to blame. It is possible that sea-ice in the ACCESS-S2 system
is left in an unbalanced state after assimilating atmospheric and
oceanic data but not sea-ice data, leading to large responses that are
amplified by the thin ice in the initial states which then subside at
longer lead times when the model is balanced.

@@

\caption{\label{fig-rmse_lon-2}Same as Figure~\ref{fig-rmse_lon-1} but
for ACCESS-S2}

@@

For ACCESS-S1 forecasts (Figure~\ref{fig-rmse_lon-1}), skill tends to be
lower off the coast of Eastern Antarctica even at short lead times; for
instance, the skill score for forecasts initialised in May and June are
negative between 0° and 120°E even at almost zero lead time. This
mirrors \citet{libera2022} findings of a ``winter predictability
barrier'', although they focus on the Weddell sea and here we show that
the effect seems to be stronger more to the East. In western Antarctica
there is a hint of easterly-propagating skill in forecasts initialised
in February and March. This is consistent with \citet{holland2013}
findings that memory of sea-ice anomalies are stored in ocean heat
content anomalies that are transported east by the Antarctic Circumpolar
Current.

ACCESS-S2 forecasts (Figure~\ref{fig-rmse_lon-2}) also have lower skill
over Eastern Antarctica. From July to December even though the
pan-Antarctic average skill is negative at all lead times
(Fig.~\ref{fig-lead-time-window}), it is positive for up to a month in
Western Antarctica. Since oceanic and atmospheric forcing is the only
source of information, this suggests that sea-ice in this region is
particularly sensitive to oceanic and atmospheric forcing and suggests a
role of the Pacific-South American mode and the Amundsen Sea Low to
shape sea-ice concentration anomalies. The fact that this is evident in
the months in which ENSO teleconnections are more important for
atmospheric circulation also suggest the influence of tropical Pacific
variability. February and March are the only two months that can be
forecasted with marginally positive skill in large regions.

@@

Sea ice forecasts from the ACCESS-S2 system show a significant low
extent bias, particularly during late summer and early autumn. This bias
is attributed to a faster and longer melt season between January and
March, and slower growth between March and April. This underestimation
during the minimum and early freezing season is a common issue in many
seasonal-to-subseasonal (S2S) systems, suggesting potential problems
either with the model's thermodynamic representation or with short wave
radiation forcing, as shown in other climate models
\citep{zampieri2019, roach2020}. Even though ACCESS-S2 shares the same
model formulation with ACCESS-S1, the latter does not suffer from this
bias, indicating that assimilating sea-ice concentrations successfully
corrects for the negative bias that exists in the free-running model.

Ensemble spread grows quickly even when perturbations are only
implemented in the atmosphere component (in ACCESS-S1), indicating that
sea-ice is indeed responding quickly to atmospheric perturbations.
However, our analysis suggests that the atmosphere and ocean data
assimilation implemented in ACCESS-S2 is only effectively influencing
sea ice initial conditions from June to October, while the rest of the
year, the sea-ice component runs virtually free, reverting to its biased
equilibrium state.

@@

\caption{\label{fig-bias-1-osi}Ensemble mean difference between monthly
sea-ice concentration of ACCESS-S2 ensemble mean forecast at 0-month
lead time (monthly mean values forecasted from the forecast initialised
at the first of the month) and observations.}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-mathphys-num,
  referee,lineno]{sn-jnl}

@@

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew C. Wheeler; Phillip Reid},
  pdfkeywords={sea ice, seasonal predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Evaluating the Importance of Initial Conditions for Antarctic Sea
Ice Seasonal Predictability with a Fully Coupled Forecast
Model]{Evaluating the Importance of Initial Conditions for Antarctic Sea
Ice Seasonal Predictability with a Fully Coupled Forecast Model}

% author setup
\author*[1,2]{\fnm{Elio} \sur{Campitelli}}\email{elio.campitelli@monash.edu}\author[1,2]{\fnm{Ariaan} \sur{Purich}}\author[1,2]{\fnm{Julie} \sur{Arblaster}}\author[3]{\fnm{Eun-Pa} \sur{Lim}}\author[3]{\fnm{Matthew C.} \sur{Wheeler}}\author[3]{\fnm{Phillip} \sur{Reid}}
% affil setup
\affil[1]{\orgdiv{School of Earth, Atmosphere and
Environment}, \orgname{Monash University, Australia,}}
\affil[2]{\orgdiv{Securing Antarctica's Environmental
Future}, \orgname{Monash University, Australia}}
\affil[3]{\orgdiv{Research, Bureau of Meteorology, Australia}}

@@

\abstract{Accurate Antarctic sea-ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. The two
versions of the Australian Bureau of Meteorology's ACCESS seasonal
forecast system, ACCESS-S1 and ACCESS-S2, use identical model
configuration and differ only in their initial conditions; primarily in
that ACCESS-S2 does not assimilate sea-ice observations, whereas
ACCESS-S1 does.\\
This provides an convenient opportunistic experiment to assess the role
of initial conditions on Antarctic sea-ice forecasts using more than 20
years of fully coupled simulations with two 9-member ensembles. Our
analysis reveals that both systems experience an extended melt season
and delayed growth phase compared with observations. This leads to a
significant negative sea-ice extent bias, which is corrected only in
ACCESS-S1 by the data assimilation system. The impact of the differing
initial conditions on forecast errors varies dramatically by season:
summer and autumn initial conditions (January-April) provide predictive
skill for up to three months, with February initial conditions being
particularly crucial. In contrast, winter forecasts of the two systems
are statistically indistinguishable after only two weeks. Regional
analysis of forecast skill suggests that this winter predictability
barrier is most dramatic over East Antarctica, where even ACCESS-S1
shows negative skill. These findings highlight the critical importance
of comprehensive year-round sampling in predictability studies and
suggest that operational sea-ice data assimilation efforts should
prioritize the summer-autumn period when initial conditions have maximum
impact on forecast skill.}

% keywords
\keywords{sea ice,  seasonal predictability,  initial
conditions,  forecasting}

@@

However, progress in Antarctic sea-ice forecasting system has lagged
behind Arctic sea-ice forecasts due to model biases, and inherent large
variability and complexity \citep{zampieri2019, gao2024a}. Dynamical
seasonal forecasts of summer Antarctic sea ice have been shown to
perform worse than relatively simpler statistical methods
\citep{massonnet2023} and machine learning approaches (e.g.
\citet{dong2024}, \citet{lin2025}), which also underscores the need for
better understanding and physical modelling of sea-ice dynamics, and
drivers of its variability.

Good initial conditions are generally required for a good forecast,
however, it is not entirely known to what extent accurate sea-ice
initial conditions affect the quality of the forecast and at what
timescales. Exploring seasonal predictions of Arctic sea ice,
\citet{guemas2016} found that sea-ice initial conditions are important
in autumn to predict summer sea ice, but the impact wasn't as dramatic
when predicting winter sea ice. \citet{day2014} also found
seasonally-varying differences in the effect of initialisation, noting
that accurate Arctic sea-ice thickness leads to improved sea-ice
forecasts initialised in July but not when initialised in January.

For the Antarctic, \citet{holland2013} studied the initial-value
predictability of Antarctic sea ice in a perfect model study using the
CCSM3 model. They found that sea-ice and ocean initial conditions
provide predictive information to forecast sea-ice edge location several
months in advance and that some predictability is retained for up to two
years thanks to ocean heat content anomalies that are advected eastward.
This is in contrast with \citet{marchi2020}, who ran perfect model
experiments to argue that uncertainty in the predicted atmospheric state
and evolution is the main driver of uncertainty in Antarctic sea-ice
extent prediction on seasonal timescales, with sea-ice and ocean initial
conditions having lesser importance. More recently, \citet{morioka2022}
studied decadal forecasts of Antarctic sea ice and found that
initialising ocean and sea ice improved the correlation between
simulated and observed sea-ice concentration evolution in the
Amundsen--Bellingshausen Sea. It is hard to compare these studies since
they are based on forecasts initialised at different times of the year
and different frameworks: \citet{holland2013} ran 20 ensemble members
initialised on the 1st of January of a particular year,
\citet{marchi2020} ran forecasts from the 1st of March and 1st of
September, and \citet{morioka2022} ran forecasts only from the 1st of
March. \citet{marchi2020} also used a coupled ocean--sea-ice model
instead of a fully coupled model like \citet{holland2013} did.
\citet{morioka2022} used observed sea-ice initial conditions and
compared with observations, while \citet{marchi2020} and
\citet{holland2013} were perfect model studies.

@@

Since model configuration is identical between ACCESS-S1 and ACCESS-S2,
they form a sort of ``opportunistic experiment'' where the same
forecasting model was run over a long period of time with multiple
ensemble forecasts initialised throughout the year, with the only
difference being the initial conditions. This provides an opportunity to
test the effect of sea-ice initial conditions on the forecast of sea-ice
concentrations and the climate.

In this study we compare sea-ice hindcasts produced by ACCESS-S1 and
ACCESS-S2. We focus on seasonality of errors and biases and the effect
of the data assimilation system. This comparison will inform future work
with the prediction system as a research tool to better understand the
dynamics and variability of the Antarctic sea ice and its impacts on the
climate system as well as to explore the potential of using its sea-ice
forecasts for decision-making. The work will also serve as a benchmark
for future prediction systems to attempt to improve upon.

@@

Both systems take atmospheric initial conditions derived from
ERA-interim \citep{dee2011} for their hindcasts. The main difference
between the hindcasts of the two systems are the ocean and sea-ice
initial conditions. ACCESS-S1's ocean and sea-ice initial conditions
come from the Met Office FOAM system, which uses a multivariate,
incremental three-dimensional variational (3D-Var),
first-guess-at-appropriate-time (FGAT) data assimilation scheme
\citep{waters2015} and assimilates sea surface temperature (SST), sea
surface height (SSH), in situ temperature and salinity profiles, and
satellite observations of sea-ice concentration using the EUMETSAT
OSISAF product described in the next section. ACCESS-S2, on the other
hand, is initialised from ocean conditions generated by the BoM weakly
coupled ensemble data assimilation scheme described in \citet{wedd2022}.
This scheme uses an optimal interpolation method and assimilates
temperature and salinity profiles from EN4 \citep{good2013}. SSTs are
nudged to Reynolds OISSTv2.1 \citep{reynolds2007} in areas where SSTs
are over 0ºC and Sea Surface Salinity is weakly nudged to the World
Ocean Atlas 2013 climatology \citep{zweng2013}.

Of most relevance for this work, sea-ice concentrations are not
assimilated in ACCESS-S2. Assimilation cycles are performed daily. The
coupled model runs for 24 hours initialised from the previous cycle.
Then the restart file fields of the ocean component are used as first
guess in the data assimilation cycle and the innovations are used to
build the next ocean initial conditions for the following cycle. The
atmosphere fields from that daily integration are not used and instead
the model atmosphere is initialised using ERA-Interim. The sea-ice
initial conditions for the next cycle are the unaltered output of the
previous daily integration. Then the cycle starts again and the coupled
model runs for another 24 hours. During this integration the sea-ice
component is affected by the ocean innovations and the new atmosphere
initial conditions via the coupler.

The ACCESS-S1 hindcast set is made up of nine members created by
perturbing the atmospheric fields only with a random field perturbation
\citep{hudson2017} and runs for 217 days for the period 1990--2012
initialised at the first of every month. The ACCESS-S2 hindcast set used
in this study runs for the period 1981--2018. Each hindcast consists of
nine ensemble members built from three consecutive three-member
forecasts created in the same manner as ACCESS-S1 members initialised at
the first of every month and the two previous days and run for 279 days.
We analyse the ensemble mean hindcasts unless otherwise specified.

@@

We use NOAA/NSIDC's Climate Data Record V4 {[}CDR; \citet{meier2014}{]}
as the primary sea-ice verification dataset. It takes the maximum value
of the NASA Team \citep{cavalieri1984} and NASA Bootstrap
\citep{comiso2023} sea-ice concentration products to reduce their low
concentration bias \citep{meier2014, meier2021}. Both source algorithms
use data from the Scanning Multichannel Microwave Radiometer (SMMR) on
the Nimbus-7 satellite and from the Special Sensor Microwave/Imager
(SSM/I) sensors on the Defense Meteorological Satellite Program's (DMSP)
-F8, -F11, and -F13 satellites. The data has a spatial resolution of 25
by 25 km and daily from November 1978 onwards.

The European Organisation for the Exploitation of Meteorological
Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility
{[}OSI; \citet{OSISAF}{]} based on the SSMIS sensor is another
satellite-derived sea-ice concentration product. It is based on mostly
the same sensors as the NOAA CDR but computed independently using
different algorithms. Figures prepared with this dataset are provided in
the supplementary material and do not differ significantly from the ones
prepared using CDR.

@@

For evaluation purposes, we use a series of measures. Sea-ice extent is
defined as the area of the ocean at least 15\% covered by sea-ice. This
threshold is motivated by the limitations in satellite retrieval, which
is increasingly unreliable for lower sea-ice concentrations
\citep{cavalieri1991}.

@@

\caption{\label{fig-bias-1}Ensemble mean difference between monthly
sea-ice concentration of ACCESS-S2 ensemble mean forecast at 0-month
lead time (monthly mean values forecasted from the forecast initialised
at the first of the month) and observations (CDR).}

@@

\caption{\label{fig-bias-2}Same as Figure~\ref{fig-bias-1} but for
ACCESS-S1.}

@@

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth results in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Admunsen Seas.

@@

\caption{\label{fig-extent-anom}Monthly mean sea-ice extent anomalies of
the observations (black) and forecasts from ACCESS-S1 (right column;
purple) and ACCESS-S2 (left column; green) at lead times of 0, 2, 4, and
6 months. The RMSE and correlation during the overlapping period of
ACCESS-S1 and ACCESS-S2 hindcasts (1990--2013) are shown on the top left
and bottom left of each panel respectively.}

@@

Unexpectedly, for ACCESS-S2, RMSE improves with lead time, even though
the correlation degrades with lead time. This is puzzling behaviour that
goes contrary to what is usually seen in prediction models. The
explanation seems to be the mentioned increased interannual variability.
Figure~\ref{fig-extent-sd} shows the interannual standard deviation of
monthly sea-ice extent of the forecasts as a function of lead time
compared with observations. ACCESS-S1 standard deviation lies within the
observed standard deviation regardless of lead time, while ACCESS-S2
standard deviation is more than twice that of observations at zero lead
time and only approaches the observed value at nine month lead time for
most months.

@@

As an example, Figure~\ref{fig-forecast-example} shows sea-ice
concentration anomalies (top row) and sea-ice thickness and the
difference between the two models (bottom row) for 2 May 2008
initialised one day prior; being that close to initialisation date,
these are very approximately the initial conditions. ACCESS-S1 sea-ice
concentrations anomalies are very close to observations as expected from
the system assimilating these data. ACCESS-S2 sea-ice concentration
anomalies, which are not assimilated, are not as close, but the
large-scale pattern is aligned with observations. The system simulates
large positive anomalies in the Weddell and Ross Seas and slight
negative anomalies in the Amundsen and Bellingshausen Seas. The fact
that ACCESS-S2 can simulate this pattern without assimilating sea-ice
data suggests that atmospheric and oceanic forcing were the dominant
drivers. However, the magnitude of the sea-ice anomalies is too big. It
is plausible that this is due to the thinner ice simulated by ACCESS-S2
(bottom row).

@@

The fact that ACCESS-S1 and ACCESS-S2 share the same model configuration
and that the increased variability is more extreme at short lead times
(Fig.~\ref{fig-extent-sd}) suggests that the data assimilation procedure
is partly responsible. It is possible that sea-ice in the ACCESS-S2
system is left in an unbalanced state after assimilating atmospheric and
oceanic data but not sea-ice data, leading to large responses that are
amplified by the thin ice in the initial states which then subside at
longer lead times when the model is balanced.

@@

\caption{\label{fig-rmse_lon-2}Same as Figure~\ref{fig-rmse_lon-1} but
for ACCESS-S2.}

@@

For ACCESS-S1 forecasts (Figure~\ref{fig-rmse_lon-1}), skill tends to be
lower off the coast of Eastern Antarctica even at short lead times; for
instance, the skill score for forecasts initialised in May and June are
negative between 0° and 120°E even at almost zero lead time. This
mirrors \citet{libera2022} findings of a ``winter predictability
barrier'', although they focus on the Weddell Sea and here we show that
the effect seems to be stronger more to the east. In West Antarctica
there is a hint of easterly-propagating skill in forecasts initialised
in February and March. This is consistent with \citet{holland2013}
findings that memory of sea-ice anomalies are stored in ocean heat
content anomalies that are transported east by the Antarctic Circumpolar
Current.

ACCESS-S2 forecasts (Figure~\ref{fig-rmse_lon-2}) also have lower skill
over East Antarctica. From July to December even though the
pan-Antarctic average skill is negative at all lead times
(Fig.~\ref{fig-lead-time-window}), it is positive for up to a month in
West Antarctica. Since oceanic and atmospheric forcing is the only
source of information, this suggests that sea-ice in this region is
particularly sensitive to oceanic and atmospheric forcing and suggests a
role of the Pacific-South American mode and the Amundsen Sea Low to
shape sea-ice concentration anomalies. The fact that this is evident in
the months in which El Niño--Southern Oscillation teleconnections are
more important for atmospheric circulation also suggest the influence of
tropical Pacific variability. February and March are the only two months
that can be forecasted with marginally positive skill in large regions.

@@

Sea-ice forecasts from the ACCESS-S2 system show a significant low
extent bias, particularly during late summer and early autumn. This bias
is attributed to a faster and longer melt season between January and
March, and slower growth between March and April. This underestimation
during the minimum and early freezing season is a common issue in many
seasonal-to-subseasonal (S2S) systems, suggesting potential problems
either with the model's thermodynamic representation or with short wave
radiation forcing, as shown in other climate models
\citep{zampieri2019, roach2020}. Even though ACCESS-S2 shares the same
model components as ACCESS-S1, the latter does not suffer from this
bias, indicating that assimilating sea-ice concentrations successfully
corrects for the negative bias that exists in the free-running model.

Ensemble spread grows quickly even when perturbations are only
implemented in the atmosphere component (in ACCESS-S1), indicating that
sea ice is indeed responding quickly to atmospheric perturbations.
However, our analysis suggests that the atmosphere and ocean data
assimilation implemented in ACCESS-S2 is only effectively influencing
sea-ice initial conditions from June to October, while the rest of the
year, the sea-ice component runs virtually free, reverting to its biased
equilibrium state.

@@

\caption{\label{fig-bias-1-osi}Ensemble mean difference between monthly
sea-ice concentration of ACCESS-S2 ensemble mean forecast at 0-month
lead time (monthly mean values forecasted from the forecast initialised
at the first of the month) and observations (OSI).}
